Binning Strategies and Related Loss for Binned Scatterplots of Large Data

### 2016-10-01 versionForJCGS/PaperJCGS.Rnw
Begin adding scagnostics discussion

# library(dbData)
library(devtools)
library(ggplot2)
library(gridExtra)
library(hexbin)
library(Hmisc)
library(lubridate)
library(MASS)
library(plyr)
library(RColorBrewer)
library(reshape2)
library(scales)
library(tidyr)
library(xtable)

@@

Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with that binning algorithm. This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of scenarios. Some choices may be proven optimal through analytical properties, while other are data dependent and require empirical exploration of loss from binning.  Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.

@@

### Try 2d binning with MVN data
n=100000
set.seed(31415)
unifdat <- data.frame(x= runif(n, min=0, max=100),
                      y= runif(n, min=0, max=100),
                      type="Uniform",
                      record = "Fine")
normdat <- data.frame(x= rnorm(n, mean=50, sd=11),
                     y= rnorm(n, mean=50, sd=11),
                     type="Normal",
                     record = "Fine")
expdat <- data.frame(x= rexp(n, .11),
                     y= rexp(n, .11),
                     type="Exponential",
                     record = "Fine")
datlist <- list(unifdat,normdat,expdat)
### round all data to resolution alpha
alpha <- 2
for(i in 1:3){
  datlist[[i]]$roundx <- ((datlist[[i]]$x + 1)%/%alpha)*alpha
  datlist[[i]]$roundy <- ((datlist[[i]]$y + 1)%/%alpha)*alpha
}

@@

ggsave("./figure/SimScatters.png", width=8, height=5.5, dpi=300)
@
\begin{figure}[hbtp]
\begin{center}
 \includegraphics[keepaspectratio=TRUE,width=.9\textwidth]{./figure/SimScatters.png}
	\caption[Scatterplots of fine and coarse versions of the simulated bivariate data.]{\label{fig:SimDataScatterplots} Scatterplots of fine and coarse versions of the simulated bivariate data. }
	\end{center}
\end{figure}
%
%Density changes are impossible to make out in the solid black areas of the plots. 97.4\%, 98.7\% and 99.1\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.
%
The \textit{data resolution} is defined as the smallest increment between successive data values. To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units in each dimension. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution at which data is recorded.

>>>>

# library(dbData)
library(devtools)
library(ggplot2)
library(gridExtra)
library(hexbin)
library(Hmisc)
library(lubridate)
library(MASS)
library(plyr)
library(RColorBrewer)
library(reshape2)
library(scales)
library(tidyr)
library(xtable)
library(scagnostics)

@@

Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with that binning algorithm. This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of distributional scenarios. Some choices may be proven optimal through analytical properties, while other are data dependent and require empirical exploration of loss from binning. Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.

@@

### Try 2d binning with MVN data
n=100000
set.seed(31415)
unifdat <- data.frame(x= runif(n, min=0, max=100),
                      y= runif(n, min=0, max=100),
                      type="Uniform",
                      record = "Fine")
normdat <- data.frame(x= rnorm(n, mean=50, sd=11),
                     y= rnorm(n, mean=50, sd=11),
                     type="Normal",
                     record = "Fine")
expdat <- data.frame(x= rexp(n, .11),
                     y= rexp(n, .11),
                     type="Exponential",
                     record = "Fine")
datlist <- list(unifdat,normdat,expdat)
### round all data to resolution alpha
alpha <- 5
for(i in 1:3){
  datlist[[i]]$roundx <- ((datlist[[i]]$x + 1)%/%alpha)*alpha
  datlist[[i]]$roundy <- ((datlist[[i]]$y + 1)%/%alpha)*alpha
}

@@

ggsave("./figure/SimScatters.png", width=8, height=5.5, dpi=300)
@
\begin{figure}[hbtp]
\begin{center}
 \includegraphics[keepaspectratio=TRUE,width=.9\textwidth]{./figure/SimScatters.png}
	\caption[Scatterplots of fine and coarse versions of the simulated bivariate data.]{\label{fig:SimDataScatterplots} Scatterplots of fine and coarse versions of the simulated bivariate data. }
	\end{center}
\end{figure}

<<SimDataScagnostics,echo=F,include=F,eval=F>>= 
# help(package="scagnostics")
head(alldat)
scags <- data.frame(0,0,0,0,0,0,0,0,0,0,0)
counter <- 0
for (i in c("Exponential","Normal","Uniform")){
  for(j in c("Fine","Coarse")){
    counter <- counter + 1
    tmp <- filter(alldat, type==i, record=="Fine")
    print(dim(tmp))
    if(j=="Fine") scags[counter,] <- c(i,j,round(as.vector(scagnostics(tmp$x,tmp$y)),2))
    if(j=="Coarse") scags[counter,] <- c(i,j,round(as.vector(scagnostics(tmp$roundx,tmp$roundy)),2))   
    names(scags) <- c("type","record",names(scagnostics(tmp$x,tmp$y)))
  }
}
head(scags)

print(xtable(scags, digits=c(0, 0, 3,3,3,3,3,3,3,3,3,3)), include.rownames=FALSE)
@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Sat Oct 01 11:29:18 2016
\begin{table}[ht]
\ktm{
\centering
\begin{tabular}{lrrrrrrrr}
  \hline
Dist-Res & Outlie & Skewed & Clumpy & Sparse & Striate & Convex & Skinny & Stringy \\ 
  \hline
  Exp-Fine & 0.15 & 0.67 & 0.01 & 0.02 & 0.02 & 0.47 & 0.55 & 0.36 \\ 
  Exp-Coarse & 0.09 & 0.65 & 0.04 & 0.04 & 0.18 & 0.53 & 0.35 & 0.30 \\ 
  Norm-Fine & 0.07 & 0.55 & 0.00 & 0.02 & 0.01 & 0.60 & 0.31 & 0.39  \\ 
  Norm-Coarse & 0.04 & 0.65 & 0.01 & 0.04 & 0.05 & 0.62 & 0.18 & 0.31 \\ 
  Unif-Fine & 0.00 & 0.52 & 0.00 & 0.03 & 0.00  & 0.68 & 0.15 & 0.36 \\ 
  Unif-Coarse & 0.00 & 0.62 & 0.00 & 0.04 & 0.01 & 0.69 & 0.11 & 0.29 \\ 
   \hline
\end{tabular}
} % end ktm highlighting
\normalsize
\caption{\ktm{Scagnostics metrics for the six simulated datasets used in empirical exploration of spatial loss in binned scatterplots.}}
\label{tab:scagnostics}
\end{table}

%
%Density changes are impossible to make out in the solid black areas of the plots. 97.4\%, 98.7\% and 99.1\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.
%
The \textit{data resolution} is defined as the smallest increment between successive data values. To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units in each dimension. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution at which data is recorded.
